---
title: "Storm Data Analysis"
author: "Ray Qiu"
date: "October 12, 2015"
output: html_document
keep_md: true
---

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm data, and trying to answer the following questions:

* Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

* Across the United States, which types of events have the greatest economic consequences?

***

#### Data Processing

Import required libraries
```{r message = FALSE}
library(dplyr)
library(tidyr)
library(gridExtra)
library(ggplot2)
```

Read data in R
```{r}
# If the RDS file does not exist, import data from original data file
if (!file.exists("rawData.Rds")) {
    # If the data file does not exist, download it from the web.
    if (!file.exists("data.csv.bz2")) {
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "data.csv.bz2", "curl")
    }
    dat <- read.csv("data.csv.bz2", header = TRUE)
    ## Save raw data to RDS file
    saveRDS(dat, file = "rawData.Rds")
} else {
    dat <- readRDS("rawData.Rds")
}
```

##### The following addresses the first question.

Group data by EVTYPE, sum the total FATALITIES, arrange the observations in a descending order, then take the top 5.
```{r message = FALSE}
fatalities <- dat %>% 
     group_by(EVTYPE) %>%
     summarize(Total_FATAL = sum(FATALITIES)) %>%
     arrange(desc(Total_FATAL)) %>%
     top_n(10)
```

Group data by EVTYPE, sum the total INJURIES, arrange the observations in a descending order, then take the top 5.
```{r message = FALSE}
injuries <- dat %>% 
     group_by(EVTYPE) %>%
     summarize(Total_INJURIES = sum(INJURIES)) %>%
     arrange(desc(Total_INJURIES)) %>%
     top_n(10)
```

#### Results

Plot the two plots
```{r message = FALSE}
fatalities_plot <- ggplot(fatalities, aes(x = EVTYPE, y = Total_FATAL / 1000)) + 
    xlab("Event Types") +
    ylab("Fatalities (thousands)") + 
    ggtitle("Fatalities by Events") +
    theme(plot.title = element_text(lineheight=.8, face="bold")) +
    geom_bar(stat="identity") + 
    theme(text = element_text(size=8), axis.text.x = element_text(angle = 45, hjust = 1))
injuries_plot <- ggplot(injuries, aes(x = EVTYPE, y = Total_INJURIES / 1000)) + 
    xlab("Event Types") +
    ylab("Injuries (thousands)") + 
    ggtitle("Injuries by Events") +
    theme(plot.title = element_text(lineheight=.8, face="bold")) +
    geom_bar(stat="identity") +
    theme(text = element_text(size=8), axis.text.x = element_text(angle = 45, hjust = 1))
grid.arrange(fatalities_plot, injuries_plot, ncol = 2)
```

From the figure, it is easy to conclude that **TORNADO**, **EXCESSIVE HEAT**, **FLASH FLOOD**, **HEAT**, and **LIGHTNING** are the top 5 events that are the most harmful with respect to fatalities; **TORNADO**, **EXCESSIVE HEAT**, **FLOOD**, **TSTM WIND**, and **LIGHTNING** are the top 5 events that are the most harmful with respect to injuries.  In summary, **TORNADO** is the most harmful with respect to population health.

***

###### The following addresses the second question.

First, we need to prepare the data by combining both PROPDMG and PROPDMGEXP
variables to create a new variable propDamage.

```{r}
# Create a helper function
f <- function(x, e) {
    if (e == "K") {
        return(x * 10^3)
    } else if (e == "M") {
        return(x * 10^6)
    } else if (e == "B") {
        return(x * 10^9)
    } else {
        return(x)
    }
}
# Add a new column as propDamage
dat$propDamage <- mapply(f, dat$PROPDMG, dat$PROPDMGEXP)
# Add a new column as cropDamage
dat$cropDamage <- mapply(f, dat$CROPDMG, dat$CROPDMGEXP)
```

Group data by EVTYPE, sum the total propDamage, arrange the observations in a descending order, then take the top 5.
```{r message = FALSE}
properties <- dat %>% 
    group_by(EVTYPE) %>%
    summarize(Total_propDamage = sum(propDamage)) %>%
    arrange(desc(Total_propDamage)) %>%
    top_n(10)
```

Group data by EVTYPE, sum the total cropDamage, arrange the observations in a descending order, then take the top 5.
```{r message = FALSE}
crops <- dat %>% 
    group_by(EVTYPE) %>%
    summarize(Total_cropDamage = sum(cropDamage)) %>%
    arrange(desc(Total_cropDamage)) %>%
    top_n(10)
```

Put the numbers together to see the bigger picture
```{r message = FALSE}
total_damages <- full_join(properties, crops, by = "EVTYPE") %>%
    mutate(Total_propDaamge = replace(Total_propDamage, is.na(Total_propDamage), 0)) %>%
    mutate(Total_cropDamage = replace(Total_cropDamage, is.na(Total_cropDamage), 0)) %>%
    gather("DamageTypes", "Numbers", 2:3) # Convert to long format for plot
```

#### Results

Plot the combined numbers for events
```{r message = FALSE}
ggplot(total_damages, aes(x = EVTYPE, y = Numbers / 10^9, fill = DamageTypes)) + 
    xlab("Event Types") +
    ylab("Total Damages (billions)") + 
    ggtitle("Economical Impacts by Events") +
    theme(plot.title = element_text(lineheight=.8, face="bold")) +
    geom_bar(stat="identity") +
    theme(text = element_text(size=8), axis.text.x = element_text(angle = 45, hjust = 1))
```

Based on the figure, we can conclude that: **FLOOD**, **HURRICANE/TYPHOON**, and **TORNADO** are the top 3 events that have the greatest economic consequences.